---
title: "An Introduction to Statistics with R"
author: "Steve Reilly, USA Today (sreilly@usatoday.com, @BySteveReilly)"
date: "NICAR 2019, Newport Beach, CA"
output: html_document
---
Goals of the session

1. Introduction & workspace setup 
2. Using R for descriptive and summary statistics
3. Linear regression
4. Multiple regression
5. Basic visualization

Data we will use

1. Diamond prices (pre-loaded practice dataset)
2. Massachusetts school test scores (pre-loaded practice dataset)
3. California school test scores (pre-loaded practice dataset)
4. NFL game TV ratings and 2016 presidential election results, 2016-2017 (USA Today analysis)

## Load libraries
```{r, message=FALSE}
library("Ecdat")
library("ggplot2")
```

## Dataset 1: Diamond Prices

#### Load and explore data 
```{r}
data(Diamond)
```
```{r eval=FALSE}
View(Diamond)
?Diamond
```

We will use the `head` function to view the first five observations
```{r}
head(Diamond,5)
```

We can also use the `tail` function to view the observations at the end of the dataset 
```{r}
tail(Diamond)
```

The `names` function allows us to see the names of the variables in the dataset.
```{r}
names(Diamond)
```

The `str` function is a powerful way to examine the structure of a dataset. Here, we see the dataset has 220 observations of 17 variables. We can also see the format of each variable and the first few observations.
```{r}
str(Diamond)
```

The `structure` function breaks our numerical variables into quartiles and shows us the minimum, maximum, and mean. For the text variables, the output shows the number of observations for each category.
```{r}
summary(Diamond)
```

### Basic visualization with trendline
Here, we use R's base graphics system to create a simple scatterplot and add a trendline.
```{r pressure}
plot(Diamond$carat, Diamond$price, main = "Diamond Carats vs. price", xlab = 'Carat', ylab = 'Price', pch = 21, bg = 'gold', ylim = c(0,16000))

abline(lm(Diamond$price~Diamond$carat))
```

### Visualize with ggplot
R's powerful ggplot library allows us to build a more detailed visualization.
```{r}
ggplot(Diamond,aes(x=carat,y=price))+ 
  geom_point(colour="red",
             alpha = 0.2,
              size = 2) + 
  geom_smooth(method="lm") +
  labs(x= "Carat Size",
         y = "Price ($)",
         title = "Diamond Carats vs. price")
```

### Linear regresion analysis
With two lines of code, we conduct a linear regression analysis. The dependent variable (or outcome, typically on Y axis) goes first, fillowed by a `~` and the independent variable (or predictor, typically on x axis).
```{r}
lm.out <- lm(Diamond$price~Diamond$carat)
summary(lm.out)
```
#### Interpret the results

* For every one-carat increase in the diamond, its value increases $11,598.90.
* This model is statistically significant at the 99.9% confidence level (p value < 0.001). 
* There is a statistically significant relationship between the number of number of carats and price of a diamond.
* The adjusted R-squared is 0.89, meaning 89% of the variation in price is explained by carat.


## Dataset 2: Massachusetts Test Results

### Explore the data frame
```{r}
data(MCAS)
```
```{r eval=FALSE}
?MCAS
View(MCAS)
```
```{r}
head(MCAS)
tail(MCAS)
names(MCAS)
summary(MCAS) 
str(MCAS)
```

### Visualize the data 

#### Create a histogram.
Is the data normally distributed? Let's look at a histogram to see.
```{r pressure4}
a <- ggplot(MCAS,aes(totsc4)) 
a + geom_histogram(binwidth=10,color="black",fill="light blue") +
  scale_x_continuous(breaks=seq(0,50,10))+
  labs(x= "Fourth-grade test scores",
         y = "Student-teacher ratio",
         title = "Fourth-grade test scores and student-teacher ratio")
```

#### View scatterplot matrix
The scatterplot matrix helps by showing us visually which variables may be related. Here, we are selecting columns 4, 11, 12, 13, 14, and 16 and creating a grid of scatterplots comparing each variable with the other five. Which variables look like they may be related?
```{r pressure3}
plot(MCAS[,c(4,11:14,16)])
```


#### View scatterplot
It looks as though there may be a relationship between totsch4 (fourth-grade test score) and tchratio (student-teacher ratio). Let's take a closer look.
```{r pressure5}
ggplot(MCAS,aes(x=totsc4,y=tchratio))+ geom_point() + geom_smooth(method="lm") +
labs(x= "Fourth-grade test scores",
         y = "Student-teacher ratio",
         title = "Fourth-grade test scores and student-teacher ratio")
```

### Regression analysis

#### Bivariate regression: Test scores and student-teacher ratio
Here, we run a linear regression model to examine relationship between fourth-grade test scores and student-teacher ratio.
```{r}
Model1 <-lm(MCAS$totsc4~MCAS$tchratio)
summary(Model1)
```
#### Interpret the results

* A one-unit increase in the student-teacher ratio in this model decreases the average test score by 1.5 points.
* There is a statistically significant relationship between student-teacher ratio and test scores.
* The 99% confidence level says that we are 99% confident that the true value of the parameters that we are interested in lies within our confidence interval.
* The null hypothesis that the coefficient is zero can be rejected at the 1% significance level.
* However, this model only explains 6 percent of the variation in test scores.

### Linear regression: Test scores and student-teacher ratio+spending per pupil
Let's examine another variable. Here, we run a linear regression model to examine the relationship between student-teacher ratio and fourth grade test scores.
```{r}
Model2 <-lm(MCAS$totsc4~MCAS$tchratio+MCAS$regday)
summary(Model2)
```
#### Interpret the results

* This model is also statistically significant, but still only explains about 6 percent of the variation in test scores.

### Linear regression: Test scores and student-teacher ratio+spending per pupil+percent of students receiving free lunch
We can also use R's `lm` function for multiple regression. Here, we examine the strength of the relationship between student-teacher ratio,  per-capita spending per student and percent of students receiving free on fourth-grade test scores.
```{r}
Model3 <-lm(MCAS$totsc4~MCAS$tchratio+MCAS$percap+MCAS$lnchpct)
summary(Model3)
```
####Interpret the results

* There is a statistically significant relationship between all variables included in this model and test scores.
* Model 3 is a much better fit than Models 1 and 2. This model explains 67 percent of the variation in test scores.
* In this model, a one-unit increase in student-teacher ratio decreases the average test score by 0.7 points after controlling per-capita income and students eligible for free/reduced lunch.

## Example 3: California test results

### Explore the data frame

```{r}
data(Caschool)
```
```{r eval=FALSE}
View(Caschool)
```
```{r}
names(Caschool)
str(Caschool)
summary(Caschool)
head(Caschool)
```

### Visualize the data

#### Visualize individual variables
Another way to explore the data visually is to look histograms of individual variables. Let's examine at a few variables that may be of interest to us.
```{r pressure7}
hist(Caschool$testscr, main="Test Scores")
hist(Caschool$avginc, main="Average Income")
hist(Caschool$compstu, main="Computers Per Student")
hist(Caschool$str, main="Student Teacher Ratio")
```

#### View scatterplot matrix
```{r pressure6}
plot(Caschool[,c(10,11,13,14)])
```

### Regression analysis 

#### Linear regression: Test scores and student-teacher ratio
Let's examine the relationship between test scores and student-teacher ratio.
```{r}
Model4 <-lm(Caschool$testscr~Caschool$str)
summary(Model4)
```
#### Interpret the results

* There is a statistically significant relationship between student-teacher ratio and test scores.
* The null hypothesis that the coefficient is zero can be rejected at the 0.1% significance level.
* However, this model only explains roughly 5 percent of the variation in test scores.
* A one-unit increase in the student-teacher ratio in this model decreases the average test score by 2.3 points.

### Multiple regression: Test scores and student-teacher ratio + computers-per-student
Let's try and build a model which more closely predicts test scores by adding computers-per-student as a second predictor variable.
```{r}
Model5 <-lm(Caschool$testscr~Caschool$str+Caschool$compstu)
summary(Model5)
```
##### Interpret the results

* This model is becomes less statistically significant, and still only explains about 9 percent of the variation in test scores.
* After controlling for computers-per-student, in this model a one-unit increase in student-teacher-ratio equals a 0.2 point decrease in test scores. 

### Multiple regression analysis: Test scores and student-teacher ratio + computers-per-student + average income
Let's try one more regression model adding average household income as a predictor variable.
```{r}
Model6 <- lm(Caschool$testscr~Caschool$str+Caschool$compstu+Caschool$avginc)
summary(Model6)
```
#### Interpret the results

* After controlling for both computers-per-student and average income, student-teacher ratio is no longer staistically significat at th 95 percent confidence level.
* This does not mean there isn't a real-world relationship between student-teacher ratio and student test scores.
* Importantly, note that this model still only accounts for 52% of the variation in student test scores. There must be omitted variables.
* This analysis shows there is need for further inquiry. We do some reporting and analysis to find what other omitted variables need to be factored into the analysis.

## 4. NFL TV Ratings and politics

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Ratings for NFL football are way down except before game starts, when people tune in to see whether or not our country will be disrespected!</p>&mdash; Donald J. Trump (@realDonaldTrump) <a href="https://twitter.com/realDonaldTrump/status/912624892239077376?ref_src=twsrc%5Etfw">September 26, 2017</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 

#### Hypothesis test

Null hypothesis: There is no stastically significant relationship between political support for President Trump and NFL ratings.

Alternative hypothesis: There is a statistically significant relationship between political support for President Trump and NFL ratings.

####Import data
```{r}
NFL.ratings <- read.csv("https://raw.githubusercontent.com/steve-reilly/nicar19/master/nfl_ratings.csv")
```

#### View summary statistics
```{r eval=FALSE}
View(NFL.ratings)
```
```{r}
str(NFL.ratings)
summary(NFL.ratings)
```

#### Visualize data

```{r}
ggplot(NFL.ratings,aes(y=NFL.ratings$ratings.change, x=NFL.ratings$trump.percent))+ geom_point() + geom_smooth(method="lm") + 
  labs(x= "Percent of vote won by Trump (%)",
       y = "Change in NFL ratings 2016-2017",
       title = "Trump vote vs. change in NFL ratings")
```

### Linear regresion analysis: Ratings change vs. Trump Winner ###
``` {r}
lm.out <- lm(NFL.ratings$ratings.change~NFL.ratings$trump.percent)
summary(lm.out)
```
#### Interpret the results

* A one-percentage point increase in the percentage of a TV market which voted for President Trump is associated with a change of 6-unit decrease in the Sunday Night Football rating
* However, the relationship between votes for Trump and is not statistically significant at the 95% percent confidence level, so we cannot reject the null hypothesis.
* In this model, political support for President Trump explains only 8 percent of variation in Sunday Night Football TV rating changes.